Equalizer Programming Guide

### 2009-03-23 Developer/ProgrammingGuide/paper.tex
PG: Introduction to Layout API
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\usepackage{tabularx,longtable,graphicx,a4,listings,wrapfig,subfigure,textcomp}
\usepackage[absolute]{textpos}

@@

\newcommand{\tm}{\texttrademark~}
\newcommand{\rc}{\raise 1ex\hbox{{\tiny\textregistered}}~}
\newcommand{\fig}[1]{Figure~\ref{#1}}
\newcommand{\sref}[1]{Section~\ref{#1}}
\newcommand{\aref}[1]{Appendix~\ref{#1}}
\newcommand{\link}[1]{\htmladdnormallink{#1}{#1}}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
  {\caption{\small\label{fExecutionFlow}Parallel Rendering}}
  \vspace{-3ex}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/processes.pdf}
  {\caption{\small\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
The render client implements the rendering part of an application. Its
execution is passive, it has no main loop and is completely driven by
Equalizer, based on the rendering tasks received from the server. The
tasks are executed by calling the appropriate task methods (see
\sref{sTaskMethods}) in the correct thread and context. The application
either implements the task methods with application-specific code or
uses the default methods provided by Equalizer.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eqHello.png}
  {\caption{\small\label{fHello}Hello, World!}}
\end{wrapfigure}
The \textsf{eqHello} example is a minimal application to illustrate the
basic principle of an Equalizer application: The application developer
has to implement the rendering method \textsf{Channel::frameDraw},
similar to the \textsf{glutDisplayFunc} in GLUT applications. It can be
run as a stand-alone application from the command line.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/2D.pdf}
  {\caption{\small A 2D compound}}
\end{wrapfigure}
The advantage of this mode is a low, constant IO overhead for the pixel
transfers, since only color information has to be transmitted. The upper
limit is the amount of pixel data for the destination view.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
  {\caption{\small A DPlex compound}}
\end{wrapfigure}

@@

The destination channel can also be used as a source channel. Contrary
to the other compound modes, it also has to use an output and
corresponding input frame. During rendering, the frustum is 'squeezed'
to configure the pixel decomposition. The destination channel can
therefore not be rendered in place, like with the other compound modes.
\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/pixelKernels.pdf}
  {\caption{\small\label{fPixelKernels}Example pixel kernels for a
      four-to-one pixel compound}}
\end{figure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\small\label{fModel}Simplified execution model}}
\end{wrapfigure}
All rendering threads in the configuration run asynchronously to the
application's main thread. Depending on the configuration's latency,
they can fall $n$ frames behind the last frame finished by the
application thread. A latency of one frame is usually not perceived by
the user, but can increase rendering performance substantially.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/cave.pdf}
  {\caption{\small\label{fConfig}An example configuration}}
\end{figure}

@@

@@

\subsubsection{\label{sFrustum}Frustum}
\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
  {\caption{\small\label{fFrusta}Wall and projection parameters}}
  \vspace{-3ex}
\end{wrapfigure}
Compounds have a frustum description to define the physical layout of
the display environment.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/uml}
  {\caption{\small\label{fUml}UML diagram of significant Equalizer and
      eqPly classes and task methods.}}
\end{figure}

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/syncAsync}
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{figure}
\fig{fSyncAsync} shows the execution of the rendering tasks of a 2-node
2D compound without latency and with a latency of one frame. The
asynchronous execution pipelines certain rendering operations and hides
imbalances in the load distribution, resulting in an improved
framerate. With \textsf{eqPly}, a speedup of 15\% has been observed on a
five-node rendering cluster when using a latency of one frame instead of
no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
  {\caption{\small\label{fModelDist}Scene Data in eqPly}}
\end{wrapfigure}
\textsf{eqPly} uses static distributed objects to distribute the model
loaded by the application. This model can be easily extended to
versioned objects to support dynamic data changes.

@@

\begin{wrapfigure}{r}{.618\textwidth}
%\begin{figure}[ht!]\center
  \includegraphics[width=.618\textwidth]{images/objects.pdf}
  {\caption{\small\label{fObjects}Scene Graph Distribution}}
\end{wrapfigure}
Applications distributing a dynamic scene graph use the frame data
instead of the init data as the entry point to their scene graph data
structure. \fig{fObjects} shows one possible implementation, where the
identifier and version of the scene graph root are transported using the
frame data. The scene graph root then serializes and de-serializes his
immediate children by transferring their identifier and current version,
similar to the static distribution done by \textsf{eqPly}.

@@

\begin{figure}[ht!]\center
%\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.9\textwidth]{images/configInit.pdf}
  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
\end{figure}
The config initialization happens in parallel, that is, all config
initialization tasks are transmitted by the server at once and their
completion is synchronized afterwards.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
  {\caption{\small\label{fOSWindow}eq::OSWindow UML class hierarchy}}
\end{wrapfigure}
All window-system specific functionality is implemented by a
specialization of \textsf{eq::OS\-Window}. The \textsf{OSWindow} class
defines the minimal interface to be implemented for a new window
system. Each \textsf{Window} uses one \textsf{OSWindow} during
execution. This separation allows an easy implementation and adaption to
another window system or application.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/DB.png}
  {\caption{\small\label{fDBDest}Destination view of a DB compound}}
\end{wrapfigure}
After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the the frame data to ensure
that all channels render a given frame using the same position.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Render Loop}}
\end{wrapfigure}
The actual rendering uses display lists or vertex buffer objects. These
OpenGL objects are allocated using the object manager. The rendering is
done by the leaf nodes, which are small enough to store the vertex
indices in a \textsf{short} value for optimal performance with VBO's.
The leaf nodes reuse the objects stored in the object manager, or create
and set up new objects if it was not yet set up. Since one object
manager is used per thread (pipe), this allows a thread-safe sharing of
the compiled display lists or VBO's across all windows of a pipe.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\small\label{fEventProcessing}Event Processing}}
\end{wrapfigure}
The purpose of the \textsf{OSWindow} processing method,
\textsf{processEvent}, is to perform window-system specific event
processing. For example, \textsf{AGLWindow::processEvent} calls
\textsf{aglUpdateContext} whenever a resize event is received. For
further, generic processing, the \textsf{Event} is passed on to
\textsf{Window::processEvent}. This \textsf{Event} no longer contains
the native event.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
\end{wrapfigure}
The glX event handler iterates over all windows of the registered pipe
to find the originating \textsf{GLXWindowIF}, which is then used to
process the event in a fashion similar to AGL. The
\textsf{GLXWindowEvent} passes the \textsf{XEvent} to
\textsf{GLXWindowIF::\-process\-Event}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/threads.pdf}
  {\caption{\small\label{fThreads}Threads within one node process}}
\end{wrapfigure}
The receiver and command thread are terminated when the application
stops network communications using \textsf{Client::exitLocal}.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/threadModels.pdf}
  {\caption{\small\label{fThreadModels}Async, draw\_sync and local\_sync
    thread synchronization models}}
\end{figure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
The local synchronization consists of \textsf{releaseFrameLocal} to
unlock the local frame, and of \textsf{waitFrameLocal} to wait for the
unlock. For the default synchronization model \textsf{sync\_draw},
Equalizer uses the task method \textsf{frameDrawFinish} which is called
on each resource after the last \textsf{Channel::frameDraw} invocation
for this frame. Consequently, \textsf{Pipe::frameDraw\-Fi\-nish} calls
\textsf{Pipe::\-release\-Frame\-Lo\-cal} to signal that it is done
drawing the current frame, and \textsf{Node::frameDrawFinish} calls
\textsf{Pipe::waitFrameLocal} for each of its pipes to block the node
thread until the current frame has been drawn.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/mainloop.pdf}
  {\caption{\small\label{fMainLoop}Synchronization of frame tasks}}
\end{figure}

@@

\begin{figure}[h!t]
  \subfigure[]{\label{fMono}
    \includegraphics[height=3.5cm]{images/mono}}\hfil
  \subfigure[]{\label{fStereo}
    \includegraphics[height=3.5cm]{images/stereo}}\hfil
  \subfigure[]{\label{fTracked}
    \includegraphics[height=3.5cm]{images/tracked}}%
  {\caption{\small\label{fImmersive}Monoscopic, Stereoscopic and Tracked
    frusta}}
\end{figure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewUpdate.pdf}
  {\caption{\small\label{fViewUpdate}Event Flow during a View Update}}
\end{wrapfigure}
When a window is resized, a \textsf{CHANNEL\_RESIZE} event is eventually
generated. If the corresponding channel has a view,
\textsf{Channel::processEvent} sends a \textsf{VIEW\_RESIZE} event to
the application. This event contains the identifier of the view.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
  \vspace{-4ex}
\end{wrapfigure}
\fig{fDirectSend} illustrates this algorithm for three channels. The
Equalizer website contains a
presentation\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}
explaining and comparing this algorithm with the binary-swap algorithm.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
\end{wrapfigure}
For output frames, the frame data's pixel viewport is the area of the
frame buffer to read back. It will transport the offset from the source
to the destination channel, that is, the frame data pixel viewport for
input frames position the pixel data on the destination. This has the
effect that a partial framebuffer readback will end up in the same place
in the destination channels.

@@

%\begin{wrapfigure}{r}{.618\textwidth}
\begin{figure}[h!t]\center
  \includegraphics[width=.9\textwidth]{images/compositor.pdf}
  {\caption{\small\label{fCompositor}Functional diagram of the compositor}}
\end{figure}
The \textsf{Compositor} class gathers a set of static functions which
implement the various compositing algorithms and low-level
optimizations. \fig{fCompositor} provides a top-down functional overview
of the various compositor functions.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\small\label{fSlices}Volume Rendering using 3D Textures}}
\end{wrapfigure}
The full power of scalable volume rendering is however in DB (sort-last)
compounds, where the full volume is divided into separate bricks. Each
of the bricks is rendered like a separate volume. For recomposition, the
\textsf{RGBA} frame buffer data resulting from these render passes then
has to be assembled correctly.

@@

\begin{figure}[h!t]
  \subfigure[]{\label{fBlendOrtho}
    \includegraphics[width=.45\textwidth]{images/b2f_ortho.pdf}
  }\hfil
  \subfigure[]{\label{fBlendPerspective}
    \includegraphics[width=.55\textwidth]{images/b2f_perspective.pdf}
  }%
  {\caption{\small\label{fBlend}Back-to-Front Com\-po\-siting for
      Orthogonal and Perspective View Frusta}}
\end{figure}

\begin{wrapfigure}{r}{.382\textwidth}
  \vspace{-3ex}\includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
  \vspace{-5ex}
\end{wrapfigure}
For recomposition, the 2D frame buffer contents are blended to form a
seamless picture. For correct blending, the frames are ordered in the
same back-to-front order as the slices used for rendering, and use the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/statistics.png}
  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
\end{wrapfigure}
Statistics are enabled per entity using an attribute hint. The hint
determines how precise the gathered statistics are. When set to
\textsf{fastest}, the per-frame clock is sampled directly when the event
occurs. When set to \textsf{nicest}, all OpenGL commands will be
finished before sampling the event. This incurs a performance penalty,
but gives more correct results. The default setting is fastest in
release builds, and nicest in debug builds.

@@

\begin{figure}[h!t]
  \includegraphics[width=\textwidth]{images/statisticsDetail.pdf}
    {\caption{\small\label{fStatisticsDetail}Annotated Detail of the
        Statistics from \fig{fStatistics}.}}
\end{figure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netNode.pdf}
  {\caption{\small\label{fNetNode}Communication between two nodes}}
\end{wrapfigure}
\fig{fNetNode} shows the communication between two nodes. The nodes in
the listening state also have a listening connection, which is not shown
in the Figure for complexity reasons.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netSession.pdf}
  {\caption{\small\label{fNetSession}Communication between two Sessions}}
\end{wrapfigure}
Sessions instances can communicate to each other directly using an
\textsf{eq::net::Ses\-sion\-Packet}, which is derived from the
\textsf{NodePacket} and extends it by the session identifier. The
session ID is used by the node to dispatch the packet to the session.

>>>>

\usepackage{tabularx,longtable,graphicx,a4,listings,wrapfig,subfigure,textcomp,ccaption}
\usepackage[absolute]{textpos}

@@

\newcommand{\tm}{\texttrademark~}
\newcommand{\rc}{\raise 1ex\hbox{{\tiny\textregistered}}~}
\newcommand{\fig}[1]{Figure~\ref{#1}}
\newcommand{\sref}[1]{Section~\ref{#1}}
\newcommand{\aref}[1]{Appendix~\ref{#1}}
\newcommand{\link}[1]{\htmladdnormallink{#1}{#1}}
\indentcaption{2em}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
  {\caption{\label{fExecutionFlow}Parallel Rendering}}
  \vspace{-3ex}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/processes.pdf}
  {\caption{\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
The render client implements the rendering part of an application. Its
execution is passive, it has no main loop and is completely driven by
Equalizer, based on the rendering tasks received from the server. The
tasks are executed by calling the appropriate task methods (see
\sref{sTaskMethods}) in the correct thread and context. The application
either implements the task methods with application-specific code or
uses the default methods provided by Equalizer.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eqHello.png}
  {\caption{\label{fHello}Hello, World!}}
\end{wrapfigure}
The \textsf{eqHello} example is a minimal application to illustrate the
basic principle of an Equalizer application: The application developer
has to implement the rendering method \textsf{Channel::frameDraw},
similar to the \textsf{glutDisplayFunc} in GLUT applications. It can be
run as a stand-alone application from the command line.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/2D.pdf}
  {\caption{ A 2D compound}}
\end{wrapfigure}
The advantage of this mode is a low, constant IO overhead for the pixel
transfers, since only color information has to be transmitted. The upper
limit is the amount of pixel data for the destination view.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
  {\caption{ A DPlex compound}}
\end{wrapfigure}

@@

The destination channel can also be used as a source channel. Contrary
to the other compound modes, it also has to use an output and
corresponding input frame. During rendering, the frustum is 'squeezed'
to configure the pixel decomposition. The destination channel can
therefore not be rendered in place, like with the other compound modes.
\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/pixelKernels.pdf}
  {\caption{\label{fPixelKernels}Example pixel kernels for a
      four-to-one pixel compound}}
\end{figure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\label{fModel}Simplified execution model}}
\end{wrapfigure}
All rendering threads in the configuration run asynchronously to the
application's main thread. Depending on the configuration's latency,
they can fall $n$ frames behind the last frame finished by the
application thread. A latency of one frame is usually not perceived by
the user, but can increase rendering performance substantially.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/cave.pdf}
  {\caption{\label{fConfig}An example configuration}}
\end{figure}

@@

\subsection{Canvases}

Canvases provide a convenient way to configure projection surfaces. A
canvas uses a layout, which describes the logical views. The usage of
canvases and layouts is optional. Typically, each desktop window uses
one canvas.

A canvas represents one logical projection surface, e.g., a powerwall, a
curved screen or an immersive installation. One configuration might
drive multiple canvases, for example an immersive installation and an
operator station.

A canvas consists of one or more segments. A planar canvas typically has
a frustum description (see \sref{sFrustum}), which is inherited by the
segments. Non-planar frusta are configured using the segment
frusta. These frusta typically describe a physically-correct display
setup for Virtual Reality installations.

\subsubsection{Segments}

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/canvas.pdf}
  {\caption{\label{fCanvas}A canvas using four, non-consecutive segments}}
\end{wrapfigure}
A segment represents one output channel of the canvas, e.g., a projector
or a LCD. A segment has an output channel, which references the channel to
which the display device is connected.

A segment covers a part of its parent canvas, which is configured using
the segment viewport. The viewport is in normalized coordinates with
respect to the canvas. Segments might overlap (edge-blended projectors)
or have gaps between each other (\fig{fCanvas}\footnote{Dataset courtesy
  of VolVis distribution of SUNY Stony Brook, NY, USA.}). The viewport
is used to configure the segment's default frustum from the canvas
frustum description, and to place layout views correctly.

Passive stereo installations have one segment per eye pass, where the
segment for the left and right eye fully overlap.

\subsection{Layouts}

A layout is the grouping of logical views. It is used by one or more
canvases. For all given layouts, Equalizer creates destination channels
when the configuration file is loaded. These destination channels can be
referenced by compounds to configure scalable rendering.

Layouts can be switched at runtime by the application. Switching a
layout will activate different destination channels for rendering.

\subsubsection{Views}

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/layout.png}
  {\caption{\label{fLayout}A layout with four views}}
\end{wrapfigure}
A view is a logical view of the application data, in the sense used by
the Model-View-Controller pattern. It can be a scene, viewing mode,
viewing position, or any other representation of the application's data.

A view has a fractional viewport with respect to its layout.  A layout
is often fully covered by its views, but this is not a requirement.

Each view can have a frustum description. The view's frustum overrides
frusta specified at the canvas or segment level. This is typically used
for non-physically correct rendering, e.g., to compare two models
side-by-side on a canvas. If the view does not specify a frustum, it
will use the sub-frustum resulting from the covered area on the canvas.

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/wallLayout.jpg}
  {\caption{A display wall using one canvas and layout with six segments
    and two views.}}
\end{figure}

@@

\subsubsection{\label{sFrustum}Frustum}
\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
  {\caption{\label{fFrusta}Wall and projection parameters}}
  \vspace{-3ex}
\end{wrapfigure}
Compounds have a frustum description to define the physical layout of
the display environment.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/uml}
  {\caption{\label{fUml}UML diagram eqPly and relevant Equalizer classes.}}
\end{figure}

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/syncAsync}
  {\caption{\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{figure}
\fig{fSyncAsync} shows the execution of the rendering tasks of a 2-node
2D compound without latency and with a latency of one frame. The
asynchronous execution pipelines certain rendering operations and hides
imbalances in the load distribution, resulting in an improved
framerate. With \textsf{eqPly}, a speedup of 15\% has been observed on a
five-node rendering cluster when using a latency of one frame instead of
no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
  {\caption{\label{fModelDist}Scene Data in eqPly}}
\end{wrapfigure}
\textsf{eqPly} uses static distributed objects to distribute the model
loaded by the application. This model can be easily extended to
versioned objects to support dynamic data changes.

@@

\begin{wrapfigure}{r}{.618\textwidth}
%\begin{figure}[ht!]\center
  \includegraphics[width=.618\textwidth]{images/objects.pdf}
  {\caption{\label{fObjects}Scene Graph Distribution}}
\end{wrapfigure}
Applications distributing a dynamic scene graph use the frame data
instead of the init data as the entry point to their scene graph data
structure. \fig{fObjects} shows one possible implementation, where the
identifier and version of the scene graph root are transported using the
frame data. The scene graph root then serializes and de-serializes his
immediate children by transferring their identifier and current version,
similar to the static distribution done by \textsf{eqPly}.

@@

\begin{figure}[ht!]\center
%\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.9\textwidth]{images/configInit.pdf}
  {\caption{\label{fConfigInit}Config Initialization Sequence}}
\end{figure}
The config initialization happens in parallel, that is, all config
initialization tasks are transmitted by the server at once and their
completion is synchronized afterwards.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
  {\caption{\label{fOSWindow}eq::OSWindow UML class hierarchy}}
\end{wrapfigure}
All window-system specific functionality is implemented by a
specialization of \textsf{eq::OS\-Window}. The \textsf{OSWindow} class
defines the minimal interface to be implemented for a new window
system. Each \textsf{Window} uses one \textsf{OSWindow} during
execution. This separation allows an easy implementation and adaption to
another window system or application.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/DB.png}
  {\caption{\label{fDBDest}Destination view of a DB compound}}
\end{wrapfigure}
After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the the frame data to ensure
that all channels render a given frame using the same position.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\label{fRender}Main Render Loop}}
\end{wrapfigure}
The actual rendering uses display lists or vertex buffer objects. These
OpenGL objects are allocated using the object manager. The rendering is
done by the leaf nodes, which are small enough to store the vertex
indices in a \textsf{short} value for optimal performance with VBO's.
The leaf nodes reuse the objects stored in the object manager, or create
and set up new objects if it was not yet set up. Since one object
manager is used per thread (pipe), this allows a thread-safe sharing of
the compiled display lists or VBO's across all windows of a pipe.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\label{fEventProcessing}Event Processing}}
\end{wrapfigure}
The purpose of the \textsf{OSWindow} processing method,
\textsf{processEvent}, is to perform window-system specific event
processing. For example, \textsf{AGLWindow::processEvent} calls
\textsf{aglUpdateContext} whenever a resize event is received. For
further, generic processing, the \textsf{Event} is passed on to
\textsf{Window::processEvent}. This \textsf{Event} no longer contains
the native event.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\label{fEventUML}UML Diagram for Event Handling Classes}}
\end{wrapfigure}
The glX event handler iterates over all windows of the registered pipe
to find the originating \textsf{GLXWindowIF}, which is then used to
process the event in a fashion similar to AGL. The
\textsf{GLXWindowEvent} passes the \textsf{XEvent} to
\textsf{GLXWindowIF::\-process\-Event}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/threads.pdf}
  {\caption{\label{fThreads}Threads within one node process}}
\end{wrapfigure}
The receiver and command thread are terminated when the application
stops network communications using \textsf{Client::exitLocal}.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/threadModels.pdf}
  {\caption{\label{fThreadModels}Async, draw\_sync and local\_sync
    thread synchronization models}}
\end{figure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
The local synchronization consists of \textsf{releaseFrameLocal} to
unlock the local frame, and of \textsf{waitFrameLocal} to wait for the
unlock. For the default synchronization model \textsf{sync\_draw},
Equalizer uses the task method \textsf{frameDrawFinish} which is called
on each resource after the last \textsf{Channel::frameDraw} invocation
for this frame. Consequently, \textsf{Pipe::frameDraw\-Fi\-nish} calls
\textsf{Pipe::\-release\-Frame\-Lo\-cal} to signal that it is done
drawing the current frame, and \textsf{Node::frameDrawFinish} calls
\textsf{Pipe::waitFrameLocal} for each of its pipes to block the node
thread until the current frame has been drawn.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/mainloop.pdf}
  {\caption{\label{fMainLoop}Synchronization of frame tasks}}
\end{figure}

@@

\begin{figure}[h!t]
  \subfigure[]{\label{fMono}
    \includegraphics[height=3.5cm]{images/mono}}\hfil
  \subfigure[]{\label{fStereo}
    \includegraphics[height=3.5cm]{images/stereo}}\hfil
  \subfigure[]{\label{fTracked}
    \includegraphics[height=3.5cm]{images/tracked}}%
  {\caption{\label{fImmersive}Monoscopic, Stereoscopic and Tracked
    frusta}}
\end{figure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewUpdate.pdf}
  {\caption{\label{fViewUpdate}Event Flow during a View Update}}
\end{wrapfigure}
When a window is resized, a \textsf{CHANNEL\_RESIZE} event is eventually
generated. If the corresponding channel has a view,
\textsf{Channel::processEvent} sends a \textsf{VIEW\_RESIZE} event to
the application. This event contains the identifier of the view.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\label{fDirectSend}Parallel Direct Send Compositing}}
  \vspace{-4ex}
\end{wrapfigure}
\fig{fDirectSend} illustrates this algorithm for three channels. The
Equalizer website contains a
presentation\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}
explaining and comparing this algorithm with the binary-swap algorithm.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
  {\caption{\label{fAssembly}Hierarchy of assembly classes}}
\end{wrapfigure}
For output frames, the frame data's pixel viewport is the area of the
frame buffer to read back. It will transport the offset from the source
to the destination channel, that is, the frame data pixel viewport for
input frames position the pixel data on the destination. This has the
effect that a partial framebuffer readback will end up in the same place
in the destination channels.

@@

%\begin{wrapfigure}{r}{.618\textwidth}
\begin{figure}[h!t]\center
  \includegraphics[width=.9\textwidth]{images/compositor.pdf}
  {\caption{\label{fCompositor}Functional diagram of the compositor}}
\end{figure}
The \textsf{Compositor} class gathers a set of static functions which
implement the various compositing algorithms and low-level
optimizations. \fig{fCompositor} provides a top-down functional overview
of the various compositor functions.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}
\end{wrapfigure}
The full power of scalable volume rendering is however in DB (sort-last)
compounds, where the full volume is divided into separate bricks. Each
of the bricks is rendered like a separate volume. For recomposition, the
\textsf{RGBA} frame buffer data resulting from these render passes then
has to be assembled correctly.

@@

\begin{figure}[h!t]
  \subfigure[]{\label{fBlendOrtho}
    \includegraphics[width=.45\textwidth]{images/b2f_ortho.pdf}
  }\hfil
  \subfigure[]{\label{fBlendPerspective}
    \includegraphics[width=.55\textwidth]{images/b2f_perspective.pdf}
  }%
  {\caption{\label{fBlend}Back-to-Front Com\-po\-siting for
      Orthogonal and Perspective Frusta}}
\end{figure}

\begin{wrapfigure}{r}{.382\textwidth}
  \vspace{-3ex}\includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
  \vspace{-5ex}
\end{wrapfigure}
For recomposition, the 2D frame buffer contents are blended to form a
seamless picture. For correct blending, the frames are ordered in the
same back-to-front order as the slices used for rendering, and use the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/statistics.png}
  {\caption{\label{fStatistics}Statistics for a two-node 2D compound}}
\end{wrapfigure}
Statistics are enabled per entity using an attribute hint. The hint
determines how precise the gathered statistics are. When set to
\textsf{fastest}, the per-frame clock is sampled directly when the event
occurs. When set to \textsf{nicest}, all OpenGL commands will be
finished before sampling the event. This incurs a performance penalty,
but gives more correct results. The default setting is fastest in
release builds, and nicest in debug builds.

@@

\begin{figure}[h!t]
  \includegraphics[width=\textwidth]{images/statisticsDetail.pdf}
    {\caption{\label{fStatisticsDetail}Annotated Detail of the
        Statistics from \fig{fStatistics}.}}
\end{figure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netNode.pdf}
  {\caption{\label{fNetNode}Communication between two nodes}}
\end{wrapfigure}
\fig{fNetNode} shows the communication between two nodes. The nodes in
the listening state also have a listening connection, which is not shown
in the Figure for complexity reasons.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netSession.pdf}
  {\caption{\label{fNetSession}Communication between two Sessions}}
\end{wrapfigure}
Sessions instances can communicate to each other directly using an
\textsf{eq::net::Ses\-sion\-Packet}, which is derived from the
\textsf{NodePacket} and extends it by the session identifier. The
session ID is used by the node to dispatch the packet to the session.
